---
title: "Explore weather station data"
output: html_notebook
---

```{r}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(lubridate)
library(scales)
library(zoo)
```

Read in weather station data.

```{r}
weather <- read_csv("../data/weather.csv")
```
Add constant year date column to enable year-over-year comparison charts. There should be a more elegant way to do this, but it works.
```{r}
weather <- weather |> 
    mutate(month_day_constant = update(date, year = 2004))
```



Average max and min temperatures per day. This will allow for visualizing year-over-year trends, and longer.

This chart shows the day-to-day variability and would probably be more useful if looking at a specific month but there are just too many days in a year to look at it all together like this.
```{r}
weather |> 
    filter(!is.na(tmax_f)) |> 
    filter(year == 2023) |> 
    group_by(year, day_of_year) |> 
    summarize(daily_avg_high_temp = mean(tmax_f)) |> 
    ggplot(aes(x = day_of_year, y = daily_avg_high_temp, color = year)) +
    geom_line()
```

Next approach: rolling average.
```{r}
weather |> 
    filter(!is.na(tmax_f)) |> 
    filter(year >= 2000 & year <= 2005) |> 
    group_by(year, day_of_year) |> 
    summarize(daily_avg_high_temp = mean(tmax_f)) |>  
    mutate(rolling_avg_high_temp = rollmean(daily_avg_high_temp, k = 7, align = "right", fill = NA)) |> 
    ggplot(aes(x = day_of_year, y = rolling_avg_high_temp, color = as.factor(year))) +
    geom_line() +
    scale_color_brewer("Dark2")
```

Basic plot for year line chart showing temperatures to get app started.
This one doesn't work to plot year on the same x (where 1/1 is compared directly to 1/1 etc)
```{r}
weather |> 
    select(year, date, tmax_f, month_name_short) |> 
    filter(!is.na(tmax_f)) |> 
    filter(year == 1899) |> 
    group_by(date) |> 
    summarize(high_temp = mean(tmax_f)) |> 
    ggplot(aes(x = date, y = high_temp)) +
    geom_line() +
    scale_y_continuous(limits = c(min(weather$tmax_f, na.rm = TRUE),
                                  max(weather$tmax_f, na.rm = TRUE))) +
    scale_x_date(date_labels = "%B") +
    labs(x = "Date", y = "Daily High Temperature", title = "High Temperature by Day")
```

```{r}
weather |> 
    select(year, date, day_of_year, month_day, tmax_f, month_name_short) |> 
    filter(!is.na(tmax_f)) |> 
    filter(year == 2000 | year == 2005) |> 
    group_by(year, day_of_year) |> 
    summarize(high_temp = mean(tmax_f)) |>  
    ggplot(aes(x = day_of_year, y = high_temp, color = as.factor(year))) +
    geom_line() +
    scale_y_continuous(limits = c(min(weather$tmax_f, na.rm = TRUE),
                                  max(weather$tmax_f, na.rm = TRUE))) +
    scale_color_brewer("Dark2")
```


This is the graph that actually does what I want to show year-to-year comparisons.
```{r}
weather |> 
            select(year, month_day_constant, tmax_f) |> 
            filter(year == 2000 | year == 2005) |> 
            filter(!is.na(tmax_f)) |> 
            group_by(year, month_day_constant) |> 
            summarize(high_temp = mean(tmax_f)) |> 
            ggplot(aes(x = month_day_constant, y = high_temp, color = as.factor(year))) +
            geom_line() +
            scale_y_continuous(limits = c(min(weather$tmax_f, na.rm = TRUE),
                                          max(weather$tmax_f, na.rm = TRUE))) +
            scale_x_date(date_labels = "%B") +
            labs(x = "Month", y = "Daily High Temperature", title = "High Temperature by Day")
```

And the same for low temperatures
```{r}
        weather |> 
            select(year, month_day_constant, tmin_f) |> 
            filter(year == 2000 | year == 2005) |> 
            filter(!is.na(tmin_f)) |> 
            group_by(year, month_day_constant) |> 
            summarize(low_temp = mean(tmin_f)) |> 
            ggplot(aes(x = month_day_constant, y = low_temp, color = as.factor(year))) +
            geom_line() +
            scale_y_continuous(limits = c(min(weather$tmin_f, na.rm = TRUE),
                                          max(weather$tmin_f, na.rm = TRUE))) +
            scale_x_date(date_labels = "%B") +
            labs(x = "Month", y = "Daily Low Temperature", title = "Low Temperature by Day")
```







Write data to binary file for app
```{r}
weather |> 
  write_rds("../met_mysteries_app/data/weather.Rds")
```

Write data to csv for app (experimenting, the binary is larger than the csv from python)
```{r}
weather |> 
    write_csv("../met_mysteries_app/data/weather.csv")
```




