---
title: "Temperature Anomalies"
output: html_notebook
---

```{r}
library(tidyverse)
library(ggplot2)
library(anomalize)
library(slider)
library(ggridges)
```

Read in daily temperature data. One row per date with tmax and tmin as the mean of all station readings from that day. Stations are in close geographic proximity so by using this technique I'm avoiding the issue of missing readings for any given station and also minimizing the impact of erroneous readings. All dates have at least 3 station readings of maximum and minimum temperatures for that day.
```{r}
weather <- read_csv("../data/temps_daily.csv")
```

Add constant year date column to enable year-over-year comparisons more easily. Come back and replace this if I figure out a way to get month labels on the x-axis with just month and day in the column (I haven't been able to figure out how to get R to recognize that as a date format).
I've actually seen other people (who know what they're doing) use this exact strategy so I think this is an acceptable strategy even though it feels like a kludge.
I picked 2004 because it was a leap year - otherwise it has no significance.
```{r}
weather <- weather |> 
    mutate(month_day_constant = update(date, year = 2004))
```

The first part is recreating some graphs from the EDA notebook but with this cleaner dataset. I was grouping by year for those charts anyway so using this dataset will result in less processing power being required to run the shiny app.

Compare high temperatures by year.
```{r}
weather |> 
    select(year, month_day_constant, tmax_f) |> 
    filter(year == 2000 | year == 2005) |> 
    ggplot(aes(x = month_day_constant, y = tmax_f, color = as.factor(year))) +
    geom_line() +
    scale_y_continuous(limits = c(min(weather$tmax_f),
                                  max(weather$tmax_f))) +
    scale_x_date(date_labels = "%B") +
    labs(x = "Month", y = "Daily High Temperature", color = "Year", title = "High Temperature by Day")
```

Compare low temperatures by year.
```{r}
weather |> 
    select(year, month_day_constant, tmin_f) |> 
    filter(year == 2000 | year == 2005) |> 
    ggplot(aes(x = month_day_constant, y = tmin_f, color = as.factor(year))) +
    geom_line() +
    scale_y_continuous(limits = c(min(weather$tmin_f),
                                  max(weather$tmin_f))) +
    scale_x_date(date_labels = "%B") +
    labs(x = "Month", y = "Daily Low Temperature", color = "Year", title = "Low Temperature by Day")
```

Trying to put high and low temps on one plot but it's not working - come back to it if there's time.
```{r}
# weather |> 
#     select(year, month_day_constant, tmin_f, tmax_f) |> 
#     filter(year == 2000 | year == 2005) |> 
#     ggplot(aes(x = month_day_constant)) +
#     geom_line(y = tmin_f, color = as.factor(year)) +
#     geom_line(y = tmax_f, color = as.factor(year)) +
#     scale_y_continuous(limits = c(min(weather$tmin_f),
#                                   max(weather$tmax_f))) +
#     scale_x_date(date_labels = "%B") +
#     labs(x = "Month", y = "Daily Temperatures", color = "Year", title = "High and Low Temperatures by Day")
```


Use anomalize library - is it really going to be as easy as it says?
The answer to that is no - it's not treating years like I expect it to.
```{r}
weather |> 
    filter(year >= 2015) |> 
    select(date, tmax_f) |> 
    arrange(date) |> 
    time_decompose(tmax_f, method = "twitter", frequency = "12 months", trend = "auto", merge = TRUE) |>  
    anomalize(remainder, alpha = 0.2, max_anoms = 0.01) |> 
    time_recompose() |>  
    plot_anomalies(ncol = 3, alpha_dots = 0.5)
```
Line chart to show daily temps by year. I think this should show all years in the range selected without the problems I was having before.
Good, but number of days and the variability makes the chart less informative.
```{r}
weather |> 
    select(year, month_day_constant, tmax_diff) |> 
    filter(year >= 2000) |> 
    arrange(year, month_day_constant) |> 
    ggplot() + 
    geom_line(aes(x = month_day_constant, y = tmax_diff, group = year, color = year)) + 
    geom_line(data = filter(weather, year == 2012), aes(x = month_day_constant, y = tmax_diff), color = "red") + 
    scale_y_continuous(limits = c(min(weather$tmax_diff),
                                  max(weather$tmax_diff))) +
    scale_x_date(date_labels = "%B") +
    scale_color_viridis_c() +
    labs(x = "Month", y = "High Temperature Difference from Mean", color = "Year", title = "High Temperature Differences from Mean by Day")
```

Create sliding window with lag and lead to smooth it out a little.
Added a line for year of interest (will be a selector in shiny) and trend lines for the overall data and selected year.
```{r}
weather |> 
    select(year, date, month_day_constant, tmax_diff) |> 
    arrange(date) |> 
    mutate(window_tmax_diff = (slide_dbl(tmax_diff, ~mean(.x), .before = 3) + 
               slide_dbl(tmax_diff, ~mean(.x), .after = 3)) / 2) |> 
    filter(year >= 1990) |> 
    ggplot() +
    geom_line(aes(x = month_day_constant, y = tmax_diff, group = year, color = year)) + 
    geom_line(data = filter(weather, year == 2022), aes(x = month_day_constant, y = tmax_diff), size = 0.9, color = "sienna3") + 
    # geom_smooth(data = filter(weather, year == 2022), aes(x = month_day_constant, y = tmax_diff), method = "lm", se = FALSE, color = "red") +
    # geom_smooth(data = filter(weather, year != 2022), aes(x = month_day_constant, y = tmax_diff), method = "lm", se = FALSE, color = "blue") +
    scale_x_date(date_labels = "%B") +
    scale_color_viridis_c() +
    labs(x = "Month", y = "High Temperature Difference from Mean", color = "Year", title = "High Temperature Difference from Mean by Day")


```

Z-score based anomalies. Scores above absolute 3 (ie more than 3std away from the mean) to be considered.
Bar chart showing aggregate anomalies by month - is there a seasonal effect?
Boxplot of all z-scores by month - are there months with larger anomalies or greater variability in z-score?
```{r}
# number of anomalies by month
weather |> 
    filter(abs(tmax_zscore) >= 3) |> 
    select(month_name_short) |> 
    group_by(month_name_short) |> 
    mutate(month_name_short = factor(month_name_short, levels = month.abb)) |> 
    ggplot(aes(x = month_name_short)) +
    geom_bar(stat = "count") +
    labs(x = "Month", y = "Count of Anomalies", title = "Count of Anomalies by Decade")
```

```{r}
weather |> 
    filter(abs(tmin_zscore) >= 3)
```


```{r}
# z-score distribution by month
weather |> 
    select(month_name_short, tmax_zscore) |> 
    group_by(month_name_short) |> 
    mutate(month_name_short = factor(month_name_short, levels = month.abb)) |> 
    ggplot(aes(x = month_name_short, y = tmax_zscore)) +
    geom_boxplot() +
    labs(x = "Month", y = "High Temperature Z-scores", title = "High Temperature Z-scores by Decade")
```


Histogram showing distribution of temperatures overall
```{r}
weather |> 
    ggplot(aes(x = tmax_f)) +
    geom_histogram(binwidth = 1) +
    labs(x = "High Temperature", y = "Count of Days", title = "Distribution of High Temperatures 1899-2023")
```

Histogram of temperatures by month
```{r}
weather |> 
    mutate(month_name_short = factor(month_name_short, levels = month.abb)) |> 
    ggplot(aes(x = tmax_f)) +
    geom_histogram(binwidth = 1) +
    facet_wrap(~month_name_short, strip.position = "bottom") +
    labs(x = "High Temperature", y = "Count of Days", title = "Distribution of High Temperatures by Month")
```

Joy plot to compare temperatures for a month across decades to see if the pattern has shifted either for certain months or overall.
Month selector in shiny app.
```{r}
month_select <- weather |> 
    select(month_name_short, tmax_f, tmin_f) |> 
    filter(month_name_short == "Jul")

weather |> 
    select(month_name_short, tmax_f, decade) |> 
    filter(month_name_short == "Jul") |> 
    ggplot(aes(x = tmax_f, y = as.factor(decade), fill = stat(x))) +
    geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
    scale_fill_viridis_c() +
    coord_cartesian(xlim = quantile(month_select$tmax_f, c(0.01, 0.99))) +
    labs(x = "Maximum Temperature",
         y = "Decade",
         title = "Density Plot of High Temperatures for Selected Month by Decade")
```

Ridgeline plot for all months - want to show yearly trends.
```{r}
weather |> 
    select(tmax_f, decade) |> 
    ggplot(aes(x = tmax_f, y = as.factor(decade), fill = stat(x))) +
    geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
    scale_fill_viridis_c() +
    coord_cartesian(xlim = quantile(weather$tmax_f, c(0.01, 0.99))) +
    labs(x = "Maximum Temperature",
         y = "Decade",
         title = "Density Plot of High Temperatures by Decade")
```




Histogram of daily diff from mean temp
```{r}
weather |> 
    ggplot(aes(x = tmax_diff)) +
    geom_histogram(binwidth = 1)
```

Histogram of daily diff from mean temp by month
```{r}
weather |> 
    ggplot(aes(x = tmax_diff)) +
    geom_histogram(binwidth = 1) +
    facet_wrap(~month_name_short)
```

Line chart to show general yearly trend over time.
Option 1: do monthly average with stacked line chart (using constant year and year as color).
```{r}
weather |> 
    select(year, month_name_short, tmax_f) |> 
    group_by(year, month_name_short) |> 
    summarize(tmax_mean = mean(tmax_f)) |> 
    mutate(month_name_short = factor(month_name_short, levels = month.abb)) |>
    ggplot(aes(x = month_name_short, y = tmax_mean, group = year, color = year)) +
    geom_line()
```

Option 2: take the diff from mean monthly average and do stacked line chart.
Not quite what I was expecting.
```{r}
weather |> 
    select(year, month_name_short, tmax_diff) |> 
    group_by(year, month_name_short) |> 
    summarize(mean_tmax_diff = mean(tmax_diff)) |> 
    mutate(month_name_short = factor(month_name_short, levels = month.abb)) |> 
    ggplot(aes(x = month_name_short, y = mean_tmax_diff, group = year, color = year)) +
    geom_line()
```

Not putting this in the app, I need more time to figure out how to make this chart work.
```{r}
# Calculate monthly averages
monthly_avg_data <- weather  |> 
    mutate(month = format(date, "%Y-%m"))  |> 
    group_by(month)  |> 
    summarize(mean_temperature = mean(tmax_f, na.rm = TRUE))

# Create a line plot
ggplot(monthly_avg_data, aes(x = as.Date(paste0(month, "-01"), format="%Y-%m-%d"), y = mean_temperature)) +
    geom_line() + 
    geom_smooth(method = "lm", se = FALSE, color = "blue") + 
    labs(x = "Month", y = "Monthly Mean Temperature", title = "Monthly Mean Temperature Over Time") +
    theme_minimal()
```
```{r}
month_choices <- weather |>
    pull(month_name) |>
    unique() |>
    factor(levels = month.name) |> 
    sort()

# month_choices <- c("All", month_choices)
```


```{r}
decadez_choices <- weather |>
    # filter(between(year, 1900, 2019)) |> 
    pull(decade) |>
    unique() |>
    sort()
```













Write data to file for app
```{r}
weather |> 
    select(year, date, month_name_short, tmax_f, tmin_f, tmax_diff, tmin_diff, tmax_zscore, tmin_zscore, decade, month_day_constant) |> 
  write_csv("../met_mysteries_app/data/weather.csv")
```